---
title: "Unsupervised Learning on Spatial Tumor Data"
author: "Alex Li"
date: "June 14th, 2022"
output: html_document
---

```{r setup}
datadir <- '~/UROP'
source(file.path(datadir, 'R/unsupervised.R'))
source(file.path(datadir, 'R/analysis.R'))
```


```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Introduction

We now demonstrate our unsupervised model on a cancer-containing spatial dataset that has initial cell type labels. The initial cancer-cell labels were determined using known marker genes, while the other labels were generated with RCTD. 

## Data preprocessing

First, we add the cancer cell labels to the raw RCTD object and augment the original SpatialRNA object with the full list of genes.

```{r preprocess, eval=F}
myRCTD <- readRDS(file.path(datadir, '/Data/myRCTD_201014_03.rds'))
results <- read.csv(file = file.path(datadir, '/Data/results_processed_slideseq_data_2020-12-12_Puck_201014_03_2020-12-12_Puck_201014_03_cell_types.csv'))

cancer_results <- results[results$HRS_location == TRUE,]
results <- myRCTD@results$results_df
results$first_type = factor(results$first_type, levels=append(levels(results$first_type), 'Cancer_cells'))
results$second_type = factor(results$second_type, levels=append(levels(results$second_type), 'Cancer_cells'))
results[cancer_results$barcodes, 'spot_class'] <- 'singlet'
results[cancer_results$barcodes,'first_type'] <- 'Cancer_cells'
# below modifies cancer cell weight, although this is not actually necessary
#for (barcode in cancer_results$barcodes) {
#  if (barcode %in% rownames(myRCTD@results$weights_doublet)) {
#    myRCTD@results$weights_doublet[barcode,'first_type'] <- 1
#  }
#}

myRCTD@results$results_df <- results
myRCTD@cell_type_info$info[[2]] <- append(myRCTD@cell_type_info$info[[2]], 'Cancer_cells')
myRCTD@cell_type_info$renorm[[2]] <- append(myRCTD@cell_type_info$renorm[[2]], 'Cancer_cells')
myRCTD@cell_type_info$info[[3]] <- myRCTD@cell_type_info$info[[3]] + 1
myRCTD@cell_type_info$renorm[[3]] <- myRCTD@cell_type_info$renorm[[3]] + 1
class_df <- myRCTD@internal_vars$class_df
class_df['Cancer_cells', 'class'] = 'Cancer_cells'
myRCTD@internal_vars$class_df <- class_df
myRCTD@config$max_cores <- 4

counts <- read.csv(file = file.path(datadir, '/Data/MappedDGEForR.csv'))
rownames(counts) <- counts$GENE; counts$GENE <- NULL
nUMI <- colSums(counts)
coords <- myRCTD@originalSpatialRNA@coords
puck <- SpatialRNA(coords, counts, nUMI)
myRCTD@originalSpatialRNA <- puck
saveRDS(puck, file.path(datadir, '/Data/cancer_puck.rds'))
saveRDS(myRCTD, file.path(datadir, '/Objects/vignette_v1/cancer_RCTD_initial.rds'))
```

## Running semisupervised learning

We now run the unsupervised learning algorithm with the updated gene list and the new labels, using CSIDE on the new labels to generate our initialization.

```{r unsupervised, eval=F}
RCTD_init <- readRDS(file.path(datadir, 'Objects/vignette_v1/cancer_RCTD_initial.rds'))
RCTD_init@config$MIN_CHANGE_DE <- 0.001
#RCTD_list <- run.de.initialization(RCTD_init, convergence_thresh = 2e-3, max_iter = 50)
cancer_genes <- readRDS(file.path(datadir, 'Data/HRS_signature.rds'))
gene_list <- intersect(union(cancer_genes, RCTD_init@internal_vars$gene_list_bulk), rownames(RCTD_init@originalSpatialRNA@counts))

cti <- cell_type_info_from_de(RCTD_init, gene_list = gene_list)
myRCTD <- create.object(
  RCTD_init@originalSpatialRNA, 
  cti, 
  gene_list_reg = gene_list
)
RCTD_list <- run.algorithm(myRCTD)

saveRDS(RCTD_list, file.path(datadir, '/Objects/vignette_v1/cancer_RCTD_new.rds'))
```

## Plotting cell types

We can plot the cell types spatially before and after the learning algorithm to compare the assignments.

```{r plot_cancer}
RCTD_init <- readRDS(file.path(datadir, 'Objects/vignette_v1/cancer_RCTD_initial.rds'))
RCTD_list <- readRDS(file.path(datadir, 'Objects/vignette_v1/cancer_RCTD_new.rds'))
RCTD_pred <- RCTD_list[[length(RCTD_list)]]
plot_all_cell_types(RCTD_init@results$results_df, RCTD_init@originalSpatialRNA@coords, RCTD_init@cell_type_info$renorm[[2]], '..')
plot_all_cell_types(RCTD_pred@results$results_df, RCTD_pred@originalSpatialRNA@coords, RCTD_pred@cell_type_info$renorm[[2]], '..')
```

## Comparing with a reference

We can now verify the predicted cell types against a reference. We start by looking at marker genes.

```{r}
reference_basic <- readRDS(file.path(datadir, 'Data/data_Steidl_dataset_082221_reanalyzed_annotated_steidl_dataset_steidl_reference_basic.rds'))
reference_basic_cti <- process_cell_type_info(reference_basic, levels(reference_basic@cell_types))
puck <- readRDS(file.path(datadir, '/Data/cancer_puck.rds'))

cur_cell_types <- setdiff(levels(reference_basic@cell_types), 'Activated_cycling_T_cells')
cell_type_info_restr = reference_basic_cti
cell_type_info_restr[[1]] = cell_type_info_restr[[1]][, cur_cell_types]
cell_type_info_restr[[2]] = cur_cell_types; cell_type_info_restr[[3]] = length(cur_cell_types)
de_genes <- get_de_genes(cell_type_info_restr, puck, fc_thresh = 2, expr_thresh = .0001, MIN_OBS = 3)
marker_data_de = get_marker_data(cell_type_info_restr[[2]], cell_type_info_restr[[1]], de_genes)

genes <- intersect(rownames(marker_data_de[marker_data_de$cell_type == 'CD4_T_cells', ]), RCTD_pred@internal_vars$gene_list_reg)

normalized_info <- apply(RCTD_pred@cell_type_info$renorm[[1]][genes, ], MARGIN = 1, FUN = function(x) log(x) - mean(log(x)))
data <- melt(normalized_info)
ggplot(data, aes(factor(Var2), factor(Var1), fill= value)) +
    geom_tile() +
    theme_classic() +
    scale_fill_gradientn(colors = pals::brewer.rdbu(20)[1:20], limits=c(-15,15)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Marker Genes')+ ylab('Cell Types')
```

## Cancer marker genes

We can also find marker genes for the new cancer cell type.

```{r marker, message = T}
get_marker_data <- function(cell_type_names, cell_type_means, gene_list) {
  marker_means = cell_type_means[gene_list,]
  marker_norm = marker_means / rowSums(marker_means)
  marker_data = as.data.frame(cell_type_names[max.col(marker_means)])
  marker_data$max_epr <- apply(cell_type_means[gene_list,],1,max)
  colnames(marker_data) = c("cell_type",'max_epr')
  rownames(marker_data) = gene_list
  marker_data$log_fc <- 0
  epsilon <- 1e-9
  for(cell_type in unique(marker_data$cell_type)) {
    cur_genes <- gene_list[marker_data$cell_type == cell_type]
    other_mean = rowMeans(cell_type_means[cur_genes,cell_type_names != cell_type])
    marker_data$log_fc[marker_data$cell_type == cell_type] <- log(epsilon + cell_type_means[cur_genes,cell_type]) - log(epsilon + other_mean)
  }
  return(marker_data)
}
cur_cell_types <- c('CD4_T_cells', 'CD8_T_cells', 'Monocytes_macrophages', 'Cancer_cells')
puck <- RCTD_pred@spatialRNA
cell_type_info_restr = RCTD_pred@cell_type_info$info
cell_type_info_restr[[1]] = cell_type_info_restr[[1]][,cur_cell_types]
cell_type_info_restr[[2]] = cur_cell_types; cell_type_info_restr[[3]] = length(cur_cell_types)
de_genes <- get_de_genes(cell_type_info_restr, puck, fc_thresh = 1, expr_thresh = .0001, MIN_OBS = 3)
marker_data_de = get_marker_data(cell_type_info_restr[[2]], cell_type_info_restr[[1]], de_genes)
cancer_data_de = marker_data_de[marker_data_de$cell_type == 'Cancer_cells', ]
cancer_data_de[order(cancer_data_de$log_fc, decreasing = T),]
```

We can also observe the expression of known marker genes for cancer cells across the predicted cell types.

```{r cancer_marker}
cancer_genes <- readRDS(file.path(datadir, 'Data/HRS_signature.rds'))
cell_type_info <- RCTD_pred@cell_type_info$info[[1]]
normalized_info <- apply(cell_type_info[intersect(rownames(cell_type_info), cancer_genes), ], MARGIN = 1, FUN = function(x) log(x) - mean(log(x)))
data <- melt(normalized_info)
ggplot(data, aes(factor(Var2), factor(Var1), fill= value)) +
    geom_tile() +
    theme_classic() +
    scale_fill_gradientn(colors = pals::brewer.rdbu(20), limits=c(-10,10)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Marker Genes')+ ylab('Cell Types')
```
```{r}
cancer_genes <- readRDS(file.path(datadir, 'Data/HRS_signature.rds'))
cell_type_info <- RCTD_pred@cell_type_info$info[[1]]
normalized_info <- apply(cell_type_info[intersect(rownames(cell_type_info), cancer_genes), ], MARGIN = 1, FUN = function(x) x / max(x))
data <- melt(normalized_info)
ggplot(data, aes(factor(Var2), factor(Var1), fill= value)) +
    geom_tile() +
    theme_classic() +
    scale_fill_gradientn(colors = pals::brewer.blues(20), limits=c(0, 1)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Marker Genes')+ ylab('Cell Types')
```


## Immune cell subtypes

We can now explore immune cell subtypes by running full mode on only pixels of a single cell type. We will focus on T-cells.

```{r eval=F}
RCTD_list <- readRDS(file.path(datadir, 'Objects/vignette_v2/cancer_RCTD_final.rds'))
RCTD_pred <- RCTD_list[[length(RCTD_list)]]
```

```{r}
T_cell_RCTD <- initialize.subtypes(RCTD_pred, c('CD4_T_cells', 'CD8_T_cells'), resolution = 1.2)
T_cell <- run.subtypes(T_cell_RCTD)
saveRDS(T_cell, file.path(datadir, 'Objects/vignette_v1/T_cell.rds'))
```


### CD4 T Cells

```{r eval=F}
CD4_RCTD <- initialize.subtypes(RCTD_pred, 'CD4_T_cells', resolution = 0.3)
CD4_RCTD <- run.subtypes(CD4_RCTD)
saveRDS(CD4_RCTD, file.path(datadir, 'Objects/vignette_v1/CD4_RCTD.rds'))
```

### CD8 T Cells

```{r eval=F}
CD8_RCTD <- initialize.subtypes(RCTD_pred, 'CD8_T_cells', resolution = 0.7)
CD8_RCTD <- run.subtypes(CD8_RCTD)
saveRDS(CD8_RCTD, file.path(datadir, 'Objects/vignette_v1/CD8_RCTD.rds'))
```

We can spatially plot cell subtypes.

```{r subtype_plot, fig.height = 6, fig.width = 12}
predicted_plot <- function(x) {
  plot_puck_continuous(RCTD_pred@originalSpatialRNA, colnames(RCTD_pred@originalSpatialRNA@counts), RCTD_pred@results$weights[, x], size = 0.1, my_pal = pals::brewer.blues(20)[2:20], title = sprintf('Cluster %s', x))
} 

CD4_RCTD <- readRDS(file.path(datadir, 'Objects/vignette_v1/CD4_RCTD.rds'))
RCTD_pred <- CD4_RCTD[[length(CD4_RCTD)]]
plots = list()

for (i in 0:2) {
  plots[[i+1]] <- predicted_plot(as.character(i))
}

ggarrange(plotlist = plots)

CD8_RCTD <- readRDS(file.path(datadir, 'Objects/vignette_v1/CD8_RCTD.rds'))
RCTD_pred <- CD8_RCTD[[length(CD8_RCTD)]]
plots = list()

for (i in 0:2) {
  plots[[i+1]] <- predicted_plot(as.character(i))
}

ggarrange(plotlist = plots)
```

We can compare the unsupervised subtypes to reference datasets.

```{r}
reference_int <- readRDS(file.path(datadir, 'Data/data_Steidl_dataset_082221_reanalyzed_annotated_steidl_dataset_steidl_reference_int.rds'))
reference_int_cti <- process_cell_type_info(reference_int, c('CD4_T_helper_cells', 'Regulatory_T_cells', 'Naive_CD4_T_cells', 'Naive_CD8_T_cells', 'Cytotoxic_CD8_T_cells', 'Activated_cycling_T_cells'))
puck <- readRDS(file.path(datadir, 'Data/cancer_puck.rds'))

get_marker_data <- function(cell_type_names, cell_type_means, gene_list) {
  marker_means = cell_type_means[gene_list,]
  marker_norm = marker_means / rowSums(marker_means)
  marker_data = as.data.frame(cell_type_names[max.col(marker_means)])
  marker_data$max_epr <- apply(cell_type_means[gene_list,],1,max)
  colnames(marker_data) = c("cell_type",'max_epr')
  rownames(marker_data) = gene_list
  marker_data$log_fc <- 0
  epsilon <- 1e-9
  for(cell_type in unique(marker_data$cell_type)) {
    cur_genes <- gene_list[marker_data$cell_type == cell_type]
    other_mean = rowMeans(cell_type_means[cur_genes,cell_type_names != cell_type])
    marker_data$log_fc[marker_data$cell_type == cell_type] <- log(epsilon + cell_type_means[cur_genes,cell_type]) - log(epsilon + other_mean)
  }
  return(marker_data)
}

cur_cell_types <- c('Regulatory_T_cells', 'Naive_CD8_T_cells', 'Cytotoxic_CD8_T_cells')
cell_type_info_restr = reference_int_cti
cell_type_info_restr[[1]] = cell_type_info_restr[[1]][, cur_cell_types]
cell_type_info_restr[[2]] = cur_cell_types; cell_type_info_restr[[3]] = length(cur_cell_types)
de_genes <- get_de_genes(cell_type_info_restr, puck, fc_thresh = 1.5, expr_thresh = .0001, MIN_OBS = 3)
marker_data_de = get_marker_data(cell_type_info_restr[[2]], cell_type_info_restr[[1]], de_genes)

CD4 <- CD4_RCTD[[length(CD4_RCTD)]]
CD4@spatialRNA <- CD4@originalSpatialRNA
CD4_cti <- cell_type_info_from_de(CD4, cell_types = c('0', '1', '2'), gene_list = rownames(marker_data_de))

#tab <- marker_data_de[marker_data_de$cell_type == 'Cytotoxic_CD8_T_cells',]
#tail(rownames(tab[order(tab$log_fc),]))

CD8 <- CD8_RCTD[[length(CD8_RCTD)]]
CD8@spatialRNA <- CD8@originalSpatialRNA
CD8_cti <- cell_type_info_from_de(CD8, cell_types = c('0', '1', '2'), gene_list = rownames(marker_data_de))

genes <- rownames(marker_data_de[marker_data_de$cell_type == 'Naive_CD4_T_cells', ])
normalized_info <- apply(CD4_cti$info[[1]][genes, ], MARGIN = 1, FUN = function(x) x/max(x))
heatmap(normalized_info, Colv = NA, Rowv = NA, scale="none")

genes <- rownames(marker_data_de[marker_data_de$cell_type == 'Cytotoxic_CD8_T_cells', ])
normalized_info <- apply(CD8_cti$info[[1]][genes, ], MARGIN = 1, FUN = function(x) x/max(x))
heatmap(normalized_info, Colv = NA, Rowv = NA, scale="none")
```

We can look at gene correlation.

```{r}
CD4_gene_correlation <- cor(CD4_cti[CD4_genes, ], reference_int_cti[[1]][CD4_genes, ])
data <- melt(CD4_gene_correlation)
ggplot(data, aes(Var1, Var2, fill= value)) +
                           geom_tile() +
                           theme_classic() +
                           scale_fill_gradientn(colors = pals::brewer.rdbu(20)[2:19], limits=c(-1,1), name='Correlation') +
                           theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                           xlab('True Cell Type')+ ylab('Predicted Cell Type')


CD8_gene_correlation <- cor(CD8_cti[CD8_genes, ], reference_int_cti[[1]][CD8_genes, ])
data <- melt(CD8_gene_correlation)
ggplot(data, aes(Var1, Var2, fill= value)) +
                           geom_tile() +
                           theme_classic() +
                           scale_fill_gradientn(colors = pals::brewer.rdbu(20)[2:19], limits=c(-1,1), name='Correlation') +
                           theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                           xlab('True Cell Type')+ ylab('Predicted Cell Type')

```

