---
title: "Unsupervised Learning on Simulated Spatial Data"
author: "Alex Li"
date: "June 14th, 2022"
output: html_document
---

```{r setup}
datadir <- '~/UROP'
source(file.path(datadir, 'R/unsupervised.R'))
source(file.path(datadir, 'R/analysis.R'))
```


```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Introduction

As a proof-of-concept for our unsupervised model, we run the algorithm on a simulated cerebellum dataset. The dataset is generated using a snRNA reference dataset, where every pair of cell types exhibits a uniform distribution of mixing. Only two cell types are mixed at a time, and the default doublet mode of the algorithm is tested.

## Generating the simulated dataset

First, we generate the simulated dataset using the reference.

```{r gen_simulated, eval=F}
sub_sample_cell <- function(gene_list, raw.data, cell_index, UMI_sample) {
  sub_sample = sample(rep(rownames(raw.data),raw.data[,cell_index]),UMI_sample,replace=FALSE)
  sub_sample = table(sub_sample)[as.character(gene_list)]
  names(sub_sample) = gene_list
  sub_sample[is.na(sub_sample)] <- 0
  return(sub_sample)
}

generate_sim_puck <- function(common_cell_types, gene_list, ref, trials = 30, UMI_tot = 1000) {
  n_cell_types = length(common_cell_types)
  n_conditions = 13
  boundary = ceiling(n_conditions / 2)
  N_samples = (n_cell_types * trials * n_conditions * (n_cell_types - 1))/2 # total number of samples
  first_UMI = numeric(N_samples); first_type = character(N_samples); second_type = character(N_samples)
  # first_UMI, first_type, second_type are all vectors of length N_samples. first_type, second_type store cell types,
  # first_UMI stores UMI counts
  UMI_tot = UMI_tot; UMI_step = UMI_tot / (n_conditions-1)
  UMI_tot = round(UMI_step * (n_conditions-1)); UMI1_vec = round(0:(n_conditions-1)*UMI_step)
  # above are the UMI amounts
  beads = Matrix(0, nrow = length(gene_list), ncol = N_samples)
  rownames(beads) = gene_list; colnames(beads) = 1:N_samples
  # beads that are filled as the algorithm runs, only thing that has to change is N_samples
  firstbeads = Matrix(0, nrow = length(gene_list), ncol = N_samples)
  rownames(firstbeads) = gene_list; colnames(firstbeads) = 1:N_samples
  secbeads = Matrix(0, nrow = length(gene_list), ncol = N_samples)
  rownames(secbeads) = gene_list; colnames(secbeads) = 1:N_samples
  # firstbeads and secbeads add together to create beads
  first_index_list <- numeric(N_samples); second_index_list <- numeric(N_samples);
  names(first_index_list) <- 1:N_samples; names(second_index_list) <- 1:N_samples;
  index = 1
  nUMI = ref@nUMI
  results <- matrix(ncol = 4)
  for(i in 1:(n_cell_types-1)) {
    print(paste("Progress",i))
    for(j in (i+1):n_cell_types) {
      print(paste("ProgressSecond",j))
      type1 = common_cell_types[i]; type2 = common_cell_types[j]
      for (condition in 1:n_conditions) {
        UMI1 = UMI1_vec[condition]; UMI2 = UMI_tot - UMI1
        for(t in 1:trials) {
          first_UMI[index] = UMI1; first_type[index] = type1; second_type[index] = type2
          # sets UMI values and cell types
          firstInd = sample(intersect(which(ref@cell_types == type1) , which(nUMI > 1000)),1)
          secondInd = sample(intersect(which(ref@cell_types == type2) , which(nUMI > 1000)),1)
          # samples a reference barcode index for each cell type
          first_index_list[index] <- firstInd; second_index_list[index] <- secondInd
          # stores the barcode indices in the lists
          firstbeads[,index] = as.vector(sub_sample_cell(gene_list, ref@counts, firstInd, UMI1))
          secbeads[,index] = as.vector(sub_sample_cell(gene_list, ref@counts, secondInd, UMI2))
          beads[,index] = firstbeads[,index] + secbeads[,index]
          # samples gene counts using the UMI amounts and the barcode index from above
          results <- rbind(results, c(type1, type2, UMI1, UMI2))
          index = index + 1
        }
      }
    }
  }
  results <- results[-1,]
  results <- as.data.frame(results)
  return(list(beads = beads, results = results))
}

reference <- readRDS(file.path(datadir, '/data/reference_RCTD_vec.rds'))
cell_types <- c('Astrocytes', 'Bergmann', 'Fibroblast', 'Golgi', 'Granule', 'MLI1', 'MLI2', 'Oligodendrocytes', 'Polydendrocytes', 'Purkinje')
beads_list <- generate_sim_puck(cell_types, rownames(reference@counts), reference, trials=30)
beads <- beads_list$beads; results <- beads_list$results
puck <- SpatialRNA(NULL, beads, use_fake_coords = T)
saveRDS(puck, file.path(datadir, '/objects/vignette_v1/sim_puck.rds'))
saveRDS(results, file.path(datadir, '/objects/vignette_v1/sim_puck_truth.rds'))
```

## Running unsupervised learning

Now we run unsupervised learning on the simulated puck.

```{r unsupervised, eval=F}
puck <- readRDS(file.path(datadir, 'objects/vignette_v1/sim_puck.rds'))
cell_type_info <- initialize.clusters(puck, resolution = 0.135)
myRCTD <- create.object(puck, cell_type_info)
myRCTD <- run.algorithm(myRCTD, MIN_CHANGE = 1e-3)
saveRDS(myRCTD, file.path(datadir, 'objects/vignette_v1/sim_puck_new.rds'))
```

## Analyzing singlets

We load the data and look at the final iteration as our predicted results. We can begin by analyzing its singlet assignment accuracy.

```{r singlets}
results <- readRDS(file.path(datadir, 'objects/vignette_v1/sim_puck_truth.rds'))
colnames(results) <- c('type1','type2','UMI1','UMI2')
RCTD_list <- readRDS(file.path(datadir, 'objects/vignette_v1/sim_puck_new.rds'))
RCTD_pred <- RCTD_list[[length(RCTD_list)]]
pred_results <- RCTD_pred@results$results_df

# analyze singlets
truth_types = c()
pred_types = c()
for (i in 1:dim(results)[1]) {
  if (pred_results[i,'spot_class'] == 'singlet' & strtoi(results[i,'UMI1']) > strtoi(results[i,'UMI2'])) {
    truth_types <- append(truth_types, results[i,'type1'])
    pred_types <- append(pred_types, RCTD_pred@results$results_df[i, 'first_type'])
  } else if (pred_results[i,'spot_class'] == 'singlet' & strtoi(results[i,'UMI2']) > strtoi(results[i,'UMI1'])) {
    truth_types <- append(truth_types, results[i,'type2'])
    pred_types <- append(pred_types, RCTD_pred@results$results_df[i, 'first_type'])
  }
}
mytable <- table(truth_types, pred_types)
mytable
assignment.accuracy.plot(mytable)
```

## Analyzing Purkinje-Bergmann doublets

Next, we will analyze doublets. Since Purkinje and Bergmann cells co-localize in the cerebellum, we will analyze these doublets.

```{r doublets}
doublet_truth <- results[results$type1 == 'Bergmann' & results$type2 == 'Purkinje',]
doublet_pred <- pred_results[rownames(doublet_truth),]

deconvolved_doublets <- doublet_pred[(doublet_pred$spot_class == 'doublet_certain') &
                 ((doublet_pred$first_type == '7' & doublet_pred$second_type == '4') | (doublet_pred$first_type == '4' & doublet_pred$second_type == '7')),]
first_IDs <- rownames(deconvolved_doublets[deconvolved_doublets$first_type == '7',])
second_IDs <- rownames(deconvolved_doublets[deconvolved_doublets$second_type == '7',])

first_truth <- strtoi(doublet_truth[first_IDs, 'UMI1']) / 1000
second_truth <- strtoi(doublet_truth[second_IDs, 'UMI1']) / 1000
first_pred <- RCTD_pred@results$weights_doublet[first_IDs, 'first_type']
second_pred <- RCTD_pred@results$weights_doublet[second_IDs, 'second_type']

true_prop <- c(first_truth, second_truth)
pred_prop <- c(first_pred, second_pred)
```

We can calculate the mean squared error of spot type proportions for these doublets.

```{r mse}
sqrt(sum((true_prop - pred_prop)^2) / length(true_prop))
```

We can also plot the predicted Bergmann proportion as a function of the true Bergmann proportion.

```{r proportion_plot}
prop_df <- data.frame(factor(true_prop), pred_prop)
true_proportions <- levels(prop_df$factor.true_prop.)
means <- c()
stdevs <- c()
for (prop in true_proportions) {
  means <- append(means, mean(prop_df[prop_df$factor.true_prop. == prop,]$pred_prop))
  stdevs <- append(stdevs, sd(prop_df[prop_df$factor.true_prop. == prop,]$pred_prop))
}
library(ggplot2)
results_df <- data.frame(as.numeric(true_proportions), means, stdevs)
colnames(results_df) <- c('true', 'means', 'stdev')
my_pal = pals::coolwarm(20)

p1 <-ggplot2::ggplot(results_df, ggplot2::aes(x=true, y=means, colour = "true")) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::geom_line(ggplot2::aes(y=true,colour = "means")) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin=means-1.96*stdevs, ymax=means+1.96*stdevs), width=.05,
                         position=ggplot2::position_dodge(0.05)) + theme_classic() + xlab('True Bergmann Proportion')+ ylab('Predicted Bergmann Proportion')+ scale_color_manual(values=c(my_pal[20], my_pal[1]),labels = c("",""), name = "") + scale_y_continuous(breaks = c(0,0.5,1), limits = c(-.03,1.03))+ scale_x_continuous(breaks = c(0,0.5,1), limits = c(-.03,1.03))+ theme(legend.position = "none")
p1
```

Next, we can analyze the rate at which pixels are being classified as doublets by the unsupervised algorithm. 

```{r doublet_rate}
doublets_pred <- pred_results[pred_results$spot_class == 'doublet_certain',] #results filtered for confident doublets
doublets_true <- results[rownames(doublets_pred),] #real proportions of predicted doublets
results$prop <- factor(strtoi(pmin(results$UMI1, results$UMI2)) / 1000) #real proportinos of all doublets
doublets_true$prop <- factor(strtoi(pmin(doublets_true$UMI1, doublets_true$UMI2)) / 1000, levels=levels(results$prop))
doublet_classification_rate <- table(doublets_true$prop) / table(results$prop)

doublet_df <- as.data.frame(doublet_classification_rate)
doublet_df$std <- sqrt(doublet_classification_rate * (1 - doublet_classification_rate) / table(results$prop))
colnames(doublet_df) <- c('prop', 'value', 'std')
doublet_df$prop <- as.numeric(levels(doublet_df$prop))

my_pal = pals::coolwarm(20)
p2 <- ggplot(doublet_df, aes(prop,value)) + geom_line()  + 
  geom_errorbar(aes(ymin=value-1.96*std, ymax=value+1.96*std), width=.02,position=position_dodge(.001)) + theme_classic() + 
  geom_hline(yintercept=0, linetype="dashed", color = "grey", size=0.5) + geom_hline(yintercept=1, linetype="dashed", color = "grey", size=0.5) + xlab("UMI Proportion of Minority Cell Type") + ylab("Doublet Classification Rate") + scale_color_manual(values=c(my_pal[1], my_pal[20]),labels = c("Doublet"), name = "Class") + scale_y_continuous(breaks = c(0,0.5,1), limits = c(0,1))+ scale_x_continuous(breaks = c(0,0.25,0.5), limits = c(-.03,0.53))
p2
```

## Analyzing marker genes

We can also analyze how well the unsupervised model predicts gene expression profiles by analyzing marker gene expression. We analyze the expression of known Bergmann marker genes across all cell types.

```{r marker, fig.width = 12}
cluster_assignments <- rownames(mytable)[apply(mytable, 2, function(x) which(x == max(x)))]
names(cluster_assignments) <- colnames(mytable)

marker_data_truth <- readRDS(file.path(datadir, 'Data/marker_data_de_standard.RDS'))
cell_type_info <- RCTD_pred@cell_type_info$info[[1]]
colnames(cell_type_info) <- cluster_assignments
markers <- rownames(marker_data_truth[marker_data_truth$cell_type == 'Bergmann',])
normalized_info <- apply(cell_type_info[intersect(rownames(cell_type_info), markers), c('Bergmann','Granule','MLI2','Oligodendrocytes','Purkinje')], MARGIN = 1, FUN = function(x) log(x) - mean(log(x)))

data <- melt(normalized_info)
ggplot(data, aes(factor(Var2), factor(Var1), fill= value)) +
    geom_tile() +
    theme_classic() +
    scale_fill_gradientn(colors = pals::brewer.rdbu(20)[1:20], limits=c(-15, 15)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Marker Genes')+ ylab('Cell Types') + labs(fill='Normalized expression')
```

```{r marker2, fig.width = 12}
normalized_info <- apply(cell_type_info[intersect(rownames(cell_type_info), markers), c('Bergmann','Granule','MLI2','Oligodendrocytes','Purkinje')], MARGIN = 1, FUN = function(x) x / max(x))

data <- melt(normalized_info)
ggplot(data, aes(factor(Var2), factor(Var1), fill= value)) +
    geom_tile() +
    theme_classic() +
    scale_fill_gradientn(colors = pals::brewer.purples(20)[1:20], limits=c(0,1)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab('Marker Genes')+ ylab('Cell Types') + labs(fill='Fraction of maximum expression')
```

