---
title: "Unsupervised learning of Slide-seq data"
author: "Alex Li"
date: "May 11th, 2022"
output: html_document
---

```{r setup}
datadir <- '~/UROP'
source(file.path(datadir, 'R/algorithm.R'))
source(file.path(datadir, 'R/analysis.R'))
```


```{r, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
```

## Introduction

We now run our model on a real spatial dataset of the cerebellum from Slide-seq. We will compare these results against those generated by RCTD, since no ground-truth dataset exists.

## Running unsupervised learning

```{r unsupervised, eval=F}
puck <- readRDS(file.path(datadir, '/Data/puckCropped_cerebellum_slideseq.rds'))
RCTD_list <- run.unsupervised(puck, resolution = 0.5, convergence_thresh = 0.999)
saveRDS(RCTD_list, file.path(datadir, 'Objects/final/cerebellum_final.rds'))
```

## Spatially plotting cell types

Before labeling, we can spatially plot the predicted cell types to get an idea of the algorithm's performance.

```{r plot}
RCTD_list <- readRDS(file.path(datadir, 'Objects/final/cerebellum_final.rds'))
RCTD_pred <- RCTD_list[[length(RCTD_list)]]
plot_all_cell_types(RCTD_pred@results$results_df, RCTD_pred@originalSpatialRNA@coords, RCTD_pred@cell_type_info$renorm[[2]], '..')
```

## Comparison with RCTD

Since there exists no ground truth for this dataset, we can instead evaluate the model's performance by comparing against a supervised method such as RCTD. We can start by comparing singlets.

```{r RCTD}
RCTD_truth <- readRDS(file.path(datadir, 'data/RCTD_cerebellum_slideseq.rds'))
pred_results <- RCTD_pred@results$results_df
truth_results <- RCTD_truth@results$results_df

# merge MLI1 and MLI2 cell types
truth_results$first_type <- gsub('MLI1|MLI2', 'MLI', truth_results$first_type)
truth_results$second_type <- gsub('MLI1|MLI2', 'MLI', truth_results$second_type)

singlet_table <- function(truth_results, pred_results) {
  truth_singlet <- truth_results[truth_results$spot_class == 'singlet',]
  pred_singlet <- pred_results[pred_results$spot_class == 'singlet',]
  common_barcode <- intersect(row.names(truth_singlet), row.names(pred_singlet))
  truth_singlet <- truth_singlet[common_barcode,]
  pred_singlet <- pred_singlet[common_barcode,]
  truth_types = unlist(list(truth_singlet[,'first_type']))
  pred_types = unlist(list(pred_singlet[,'first_type']))
  return(table(truth_types, pred_types))
}
mytable <- singlet_table(truth_results, pred_results)
mytable
```

We can then compute the overall accuracy, taking doublets into account.

```{r accuracy}
cluster_assignments <- rownames(mytable)[apply(mytable, 2, function(x) which(x == max(x)))]
names(cluster_assignments) <- colnames(mytable)

# replace cluster with assigned cell type
ground_truth <- truth_results[truth_results$spot_class %in% c('singlet', 'doublet_certain', 'doublet_uncertain'),
                              c('spot_class', 'first_type', 'second_type')]
predicted <- pred_results[rownames(ground_truth),c('spot_class', 'first_type', 'second_type')]
predicted$first_type <- sapply(predicted$first_type, function(x) cluster_assignments[x])
predicted$second_type <- sapply(predicted$second_type, function(x) cluster_assignments[x])

# compute accuracy
cell_types <- RCTD_truth@cell_type_info$info[[2]]
cell_types <- cell_types[! cell_types %in% c('MLI1', 'MLI2')]
cell_types <- append(cell_types, 'MLI')
total <- rep(0, length(cell_types)); correct <- rep(0, length(cell_types))
names(total) <- cell_types; names(correct) <- cell_types

for (barcode in rownames(ground_truth)) {
  truth_row <- ground_truth[barcode,]
  pred_row <- predicted[barcode,]
  if (truth_row$spot_class == 'singlet' || truth_row$spot_class == 'doublet_uncertain') {
    truth_type <- truth_row$first_type
    if (pred_row$spot_class %in% c('singlet', 'doublet_uncertain')) {
      if (truth_type == pred_row$first_type) {
        total[truth_type] <- total[truth_type] + 1
        correct[truth_type] <- correct[truth_type] + 1
      } else {
        total[truth_type] <- total[truth_type] + 1
      }
    } else if (pred_row$spot_class == 'doublet_certain') {
      if (truth_type %in% c(pred_row$first_type, pred_row$second_type)) {
        total[truth_type] <- total[truth_type] + 1
        correct[truth_type] <- correct[truth_type] + 0.5
      } else {
        total[truth_type] <- total[truth_type] + 1
      }
    }
  } else if (truth_row$spot_class == 'doublet_certain') {
    truth_type <- c(truth_row$first_type, truth_row$second_type)
    if (pred_row$spot_class %in% c('singlet', 'doublet_uncertain')) {
      if (pred_row$first_type %in% truth_type) {
        total[pred_row$first_type] <- total[pred_row$first_type] + 1
        correct[pred_row$first_type] <- correct[pred_row$first_type] + 1
      } else {
        total[pred_row$first_type] <- total[pred_row$first_type] + 0.5
        total[pred_row$second_type] <- total[pred_row$second_type] + 0.5
      }
    } else if (pred_row$spot_class == 'doublet_certain') {
      if (pred_row$first_type %in% truth_type) {
        total[pred_row$first_type] <- total[pred_row$first_type] + 1
        correct[pred_row$first_type] <- correct[pred_row$first_type] + 1
      } else {
        total[pred_row$first_type] <- total[pred_row$first_type] + 1
      }
      if (pred_row$second_type %in% truth_type) {
        total[pred_row$second_type] <- total[pred_row$second_type] + 1
        correct[pred_row$second_type] <- correct[pred_row$second_type] + 1
      } else {
        total[pred_row$second_type] <- total[pred_row$second_type] + 1
      }
    }
  }
}

sum(correct) / sum(total) # overall accuracy
mydf <- as.data.frame(correct / total)
colnames(mydf) <- 'proportion'
mydf$cell_type <- factor(rownames(mydf))
mydf <- na.omit(mydf[mydf$proportion > 0, ])
p <- ggplot(mydf, aes(x=cell_type, y=proportion)) + 
    geom_point() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    xlab('Cell Type')+ ylab('Accuracy') +
    ylim(0,1)
p
```

## Marker genes

We can analyze the predicted gene expression profiles by analyzing the expression of known marker genes. We will analyze the expression of known Purkinje marker genes across the predicted cell types.

```{r marker}
marker_data_truth <- readRDS(file.path(datadir, 'Data/marker_data_de_standard.RDS'))
cell_type_info <- RCTD_pred@cell_type_info$info[[1]]
colnames(cell_type_info) <- cluster_assignments
markers <- rownames(marker_data_truth[marker_data_truth$cell_type == 'Purkinje',])
normalized_info <- apply(cell_type_info[intersect(rownames(cell_type_info), markers), ], MARGIN = 1, FUN = function(x) x/max(x))

heatmap(normalized_info, Colv = NA, Rowv = NA, scale="column")
```

